---
title: 'A Statistical Analysis of Climate in Minnesota'
author: "Caitlin Davies"
date: "8/30/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library("dplyr")
library("tidyr")
library("devtools")
library("ggplot2")
devtools::install_github("twitter/AnomalyDetection")
library("AnomalyDetection")
library("cluster")
library("scatterplot3d")
library("ggrepel")
library("knitr")
library("kableExtra")
library("formattable")
```

```{r import}

Stations <- read.csv(file = "MRCC Station Details.csv")

ADA <- read.csv("ADA_clean.csv")
ADA <- cbind(Station.Name = "ADA", ADA)

AGASSIZ_REFUGE <- read.csv("AGASSIZ_REFUGE_clean.csv")
AGASSIZ_REFUGE <- cbind(Station.Name = "AGASSIZ REFUGE", AGASSIZ_REFUGE)

ARGYLE <- read.csv("ARGYLE_clean.csv")
ARGYLE <- cbind(Station.Name = "ARGYLE", ARGYLE)

BEMIDJI_MUNICIPAL_AP <- read.csv("BEMIDJI_MUNICIPAL_AP_clean.csv")
BEMIDJI_MUNICIPAL_AP <- cbind(Station.Name = "BEMIDJI MUNICIPAL AP", BEMIDJI_MUNICIPAL_AP)

LEECH_LAKE <- read.csv("LEECH_LAKE_clean.csv")
LEECH_LAKE <- cbind(Station.Name = "LEECH LAKE", LEECH_LAKE)

POKEGAMA_DAM <- read.csv("POKEGAMA_DAM_clean.csv")
POKEGAMA_DAM <- cbind(Station.Name = "POKEGAMA DAM", POKEGAMA_DAM)

GRAND_MARAIS <- read.csv("GRAND_MARAIS_clean.csv")
GRAND_MARAIS <- cbind(Station.Name = "GRAND MARAIS", GRAND_MARAIS)

DULUTH <- read.csv("DULUTH_clean.csv")
DULUTH <- cbind(Station.Name = "DULUTH", DULUTH)

TWO_HARBORS <- read.csv("TWO_HARBORS_clean.csv")
TWO_HARBORS <- cbind(Station.Name = "TWO HARBORS", TWO_HARBORS)

ARTICHOKE_LAKE_1_E <- read.csv("ARTICHOKE_LAKE_1_E_clean.csv")
ARTICHOKE_LAKE_1_E <- cbind(Station.Name = "ARTICHOKE LAKE 1 E", ARTICHOKE_LAKE_1_E)

BEARDSLEY <- read.csv("BEARDSLEY_clean.csv")
BEARDSLEY <- cbind(Station.Name = "BEARDSLEY", BEARDSLEY)

AITKIN_2E <- read.csv("AITKIN_2E_clean.csv")
AITKIN_2E <- cbind(Station.Name = "AITKIN 2E", AITKIN_2E)

ALBERT_LEA_3_SE <- read.csv("ALBERT_LEA_3_SE_clean.csv")
ALBERT_LEA_3_SE <- cbind(Station.Name = "ALBERT LEA 3 SE", ALBERT_LEA_3_SE)

AMBOY <- read.csv("AMBOY_clean.csv")
AMBOY <- cbind(Station.Name = "AMBOY", AMBOY)

All_Stations <- rbind(ADA, AGASSIZ_REFUGE, ARGYLE, BEMIDJI_MUNICIPAL_AP, LEECH_LAKE, POKEGAMA_DAM, GRAND_MARAIS, DULUTH, TWO_HARBORS, ARTICHOKE_LAKE_1_E, BEARDSLEY, AITKIN_2E, ALBERT_LEA_3_SE, AMBOY)

```

```{r functions}

## This function allows me to convert month/day to the # day of the year, so that data across years can be easily compared to each other.

day_in_year <- function(m, d) {ifelse(m == 1, d,
    ifelse(m == 2, 31 + d,
    ifelse(m == 3, sum(31, 28) + d,
    ifelse(m == 4, sum(31, 28, 31) + d,
    ifelse(m == 5, sum(31, 28, 31, 30) + d,
    ifelse(m == 6, sum(31, 28, 31, 30, 31) + d,
    ifelse(m == 7, sum(31, 28, 31, 30, 31, 30) + d,
    ifelse(m == 8, sum(31, 28, 31, 30, 31, 30, 31) + d,
    ifelse(m == 9, sum(31, 28, 31, 30, 31, 30, 31, 31) + d,
    ifelse(m == 10, sum(31, 28, 31, 30, 31, 30, 31, 31, 30) + d,
    ifelse(m == 11, sum(31, 28, 31, 30, 31, 30, 31, 31, 30, 31) + d, 
           sum(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30) + d)))))))))))}

## This function allows me to group together winter tempuratures, so that data can be analyzed and visualized more easily.

group_cold <- function(t) {ifelse(t > 300, (t - 365), t)}
```

``` {r station_observations, message = FALSE}

Stations.Observations.Count <- All_Stations %>%
  group_by(Station.Name) %>%
  summarise(Observation.Count = n(),
            Min.Year = min(Year),
            Max.Year = max(Year))

Stations <- inner_join(Stations, Stations.Observations.Count)

```

### Size and Completeness of Collected Datasets

An imported dataset, on average, contains **`r format(mean(Stations$Observation.Count), big.mark=",")`** observations. This average hides a wide variability in dataset size:

```{r stations_boxplot, fig.width = 10}

ggplot(Stations, aes(x = factor(Climate.Division), y = Observation.Count)) +
  geom_boxplot(fill = "#4271AE", alpha = 0.7) +
  labs(x = "Climate Divisions", y = "Count of Observations", title = "Distribution of Dataset Size across Climate Division")

```

```{r Min_Max_Temp}

complete.years <- All_Stations %>% 
  select(Station.Name, Year, Month, Day) %>%
  group_by(Station.Name, Year) %>%
  summarise(
    First.Month = first(Month), 
    First.Day = first(Day),
    Last.Month = last(Month), 
    Last.Day = last(Day)
  ) %>%
  filter(First.Month == 1 & First.Day == 1 & Last.Month == 12 & Last.Day == 31)

Max.Tempurature <- All_Stations %>%
  select(Station.Name, Year, Month, Day, TMAX) %>%
  group_by(Station.Name, Year) %>%
  filter(min_rank(desc(TMAX)) <= 1) %>%
  slice(which.max(TMAX)) %>%
  semi_join(complete.years, by = c("Station.Name", "Year"))

Max.Tempurature$Day_of_Year <- day_in_year(Max.Tempurature$Month, Max.Tempurature$Day)

Min.Tempurature <- All_Stations %>%
  select(Station.Name, Year, Month, Day, TMIN) %>%
  group_by(Station.Name, Year) %>%
  slice(which.min(TMIN)) %>%
  semi_join(complete.years, by = c("Station.Name", "Year"))

Min.Tempurature$Day_of_Year <- day_in_year(Min.Tempurature$Month, Min.Tempurature$Day)

# Min.Tempurature$Day_of_Year <- group_cold(Min.Tempurature$Day_of_Year)

Min.Max.Tempurature <- full_join(Min.Tempurature, Max.Tempurature, by = c("Station.Name", "Year", "Month", "Day", "Day_of_Year"))

Min.Max.Tempurature <- Min.Max.Tempurature %>%
  gather(Extreme, Temp, c(TMIN, TMAX))

```

Some of those observations occur in partial years. (I.e., the data collected for some years does not begin on January 1st, does not end on December 31st, and/or is missing data from the middle of the year.)

```{r partial_years}

All_Stations_Complete_Years_Count <- All_Stations %>% 
  semi_join(complete.years, by = c("Station.Name", "Year"))

All_Stations_Complete_Years_Count <- All_Stations_Complete_Years_Count %>%
  select(Station.Name) %>%
  group_by(Station.Name) %>%
  summarise(Count = n())

All_Stations_Count <- All_Stations %>%
  select(Station.Name) %>%
  group_by(Station.Name) %>%
  summarise(Count = n())

All_Stations_Count_Both <- full_join(All_Stations_Count, All_Stations_Complete_Years_Count, by = "Station.Name")

colnames(All_Stations_Count_Both)[2] <- "All_Years"

colnames(All_Stations_Count_Both)[3] <- "Full_Years"

All_Stations_Count_Both$Difference <- All_Stations_Count_Both$All_Years - All_Stations_Count_Both$Full_Years

All_Stations_Count_Both$All_Years <- NULL

All_Stations_Count_Both <- gather(All_Stations_Count_Both, "Count_Type", "Count", Full_Years, Difference)

```

The imported datasets, on average, contain **`r format(mean(All_Stations_Count$Count), big.mark=",")`** observations from complete years, which is (on average) **`r percent(mean(All_Stations_Complete_Years_Count$Count) / mean(All_Stations_Count$Count))`** of the total observations. 

```{r partial_years_graph, fig.width = 10}

ggplot(All_Stations_Count_Both, aes(Station.Name, Count, fill = Count_Type)) +
  geom_col() +
  labs(x = "", y = "Observations", title = "Observations by Complete or Partial Year")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(legend.title=element_blank()) +
  scale_fill_discrete(labels=c("Partial Years", "Complete Years"))

```

Only data from complete years is used in this analysis.

----

## Observing Tempurature Extremes

----

### The Highest Tempurature of the Year

The highest tempurature of the year, across the 14 Climate Stations, is typically banded between 83 degrees and 105 degrees.

Outliers as low as 61 degrees, and as high as 115 degrees, are observed:

```{r max_temp_boxplot, fig.width = 10}

ggplot(Max.Tempurature, aes(x = Station.Name, y = TMAX)) +
  geom_boxplot(fill = "#4271AE", alpha = 0.7) +
  labs(x = "", y = "Max Tempurature", title = "Spread of Maximum Tempurature Per Year") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Days over 100 degrees have been observed throughout the datasets, often clustering together in observed heatwaves. A particularly large heatwave is observed between 1920 and 1950: 

```{r count_over_100_degrees, fig.width = 10}

Days_Over_100 <- All_Stations %>%
  select(Station.Name, Year, Month, TMAX) %>%
  group_by(Station.Name) %>%
  filter(TMAX >=100) %>%
  semi_join(complete.years, by = c("Station.Name", "Year"))

ggplot(Days_Over_100, aes(Year)) +
  geom_bar() +
  labs(y = "Count of Days Over 100 Degrees")

```

### The Hottest Day of the Year

The hottest day of the year typically occurs between June 12th and August 18th.

Outliers as early as March 12th, and as late as October 20th, are observed:

```{r max_year_boxplot, fig.width = 10}

ggplot(Max.Tempurature, aes(x = Station.Name, y = Day_of_Year)) +
  geom_boxplot(fill = "#4271AE", alpha = 0.7) +
  labs(x = "", y = "Hottest Day of the Year", title = "Visual of the Spread of Hottest Day of the Year") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r max_year_trend, fig.width = 10}

ggplot(Max.Tempurature, aes(x = Year, y = Day_of_Year, color = TMAX)) +
  geom_point() +
  geom_smooth(method = "loess", color = "black") +
  facet_wrap("Station.Name") +
  theme(legend.position='none')

```

### The Lowest Tempurature of the Year

The lowest tempurature of the year, throughout Minnesota, is typically banded between -40 degrees and -20 degrees.

Outliers as low as -60 degrees, and as high as 0 degrees, are observed:

```{r min_boxplot, fig.width = 10}

ggplot(Min.Tempurature, aes(x = Station.Name, y = TMIN)) +
  geom_boxplot(fill = "#4271AE", alpha = 0.7) +
  labs(x = "", y = "Coldest Day of the Year", title = "Visual of the Spread of Coldest Day of the Year") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


```{r min_point, fig.width = 10}
ggplot(Min.Tempurature, aes(x = Year, y = TMIN, color = Station.Name)) +
  geom_point() +
  geom_smooth(method = "loess", color = "black") +
  facet_wrap("Station.Name") +
  theme(legend.position='none')

```

```{r min_year, fig.width = 10}

ggplot(Min.Tempurature, aes(x = Year, y = Day_of_Year, color = Station.Name)) +
  geom_point() +
  geom_smooth(method = "loess", color = "black") +
  facet_wrap("Station.Name") +
  theme(legend.position='none')

```
